---
title: "Coronary Heart Disease and Stroke Mortality Outline"
output: html_document
date: "2022-11-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)

```
>Group 317B

## U.S. Coronary Heart Disease and Stroke Mortality Outline

# Introduction
Some of the leading causes of death in the United States are strokes and heart diseases. Medical data analysis is necessary to study in order to figure out how to better treat patients, which is necessary to save lives. In this project we are analyzing a collection of data about heart disease and stroke deaths in order to analyze roots and causes. This data is sourced directly from the CDC, which in turn is sourced by state governments. Through this analysis we will examine factors including sex, age, race, and state location in order to evaluate how they affect the prevalence of strokes and heart diseases.

# Background

Our question is whether age, race, and gender influence stroke and heart disease mortality, and whether geographical location influences the prevalence of these cardiovascular diseases.

Coronary heart disease is the leading cause of death in the United States for both men and women of all racial and ethnic groups, and a stroke occurs every 40 seconds. As a result, analyzing cardiovascular disease is common for many Americans and can affect anyone. Many organizations study and research both the effects and the causes in order to prevent the impact on society from continuing, but there is a general lack of funding for many health research facilities. The Division for Heart Disease and Stroke Prevention, Million Hearts, and WISEWOMEN are a few of these organizations.

Cardiovascular disease encompasses a wide range of heart conditions such as coronary artery disease, heart attack, arrhythmia, heart failure, high blood pressure, and others. Many of the symptoms of heart disease are not noticed by the patient until a major event occurs. Diabetes, obesity, an unhealthy diet, and lack of physical activity all increase one's risk of heart disease, but it can also be inherited.

# Data

Our project will use the provided data to examine the relationships between stroke and coronary heart disease, as well as patient characteristics and geographic location. This data set was collected by the National Vital Statistics System, a division of the Centers for Disease Control and Prevention, and can be found directly on the [CDC website](https://chronicdata.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Heart-Disease-and-Stroke-Morta/7b9s-s8ck). This Federal Government agency obtains its information from state medical records. The data set includes population estimates of stroke and coronary heart disease mortality rates from 1999 to 2019. The data set used a Bayesian spatiotemporal model that was smoothed over time, space, and age group to collect death rate trends.


There are over 5 million rows and 21 columns in the raw data set. We will be interested in eight variables, six of which are categorical and two of which are numerical and contain the estimated death rate value and the year data was collected. The four categorical variables are State, County, cardiovascular disease, age range, race, and gender. The state and county will be used to determine the geographical dependencies of the data values, as well as which counties have a higher prevalence. The age range indicates where the data was collected from females and males aged 35 to 64, as well as those aged 65 and older. There are also two categorical variables that contain the data's race and gender. The numeric value represents the death rate per 100,000 people from heart disease, stroke, or heart failure, as specified by the Topic column.

# Methods

```{r, include = FALSE, warning = FALSE, echo = FALSE}
#Age
all_data = read_csv("data.csv")

data = all_data %>% 
  select(Year,LocationAbbr, LocationDesc, Topic, Data_Value, Stratification1, Stratification2, Stratification3) %>% 
  rename(year = Year, state = LocationAbbr, county = LocationDesc, age_group = Stratification1, data = Data_Value, type = Topic, race = Stratification2, sex = Stratification3) %>% 
  filter(!is.na(data)) %>% 
  arrange(year)

age_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease") %>% 
  filter(race == "Overall", (sex == "Overall")) %>% 
  select(year, state, county, type, data, age_group)
age_data

age_sum = age_data %>% 
  group_by(age_group, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data)) %>% 
  distinct(total, c,n) %>% 
  mutate(p = total/n) 

#Sex
sex_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease", race == "Overall", sex != "Overall") %>% 
  select(!race)
sex_data

sex_sum = sex_data %>% 
  group_by(sex, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data)) %>% 
  distinct(total, c,n) %>% 
  mutate(p = total/n)

#Race
race_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease", race != "Overall", sex == "Overall") %>% 
  select(!sex)

race_sum = race_data %>% 
  group_by(race, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data),
    mean = mean(data)) %>% 
  distinct(total, c,n, mean) %>% 
  mutate(p = total/n)
race_sum
```

1. For the first question of interest, we will be analyzing the influence that race, sex and age have on both stroke and coronary heart disease deaths in the United States. We will be using two sample mean hypothesis testing and comparing between the two groups of both heart disease and stroke separately. For example, we will test the mean mortality rate over the years from those who died of heart disease aged 35 to 64 compared to those under similar conditions but aged 65 and older. For the difference in race, we will test against the mean of all the races combined to a set of one or two races to test if there is an influence or trend. 

Additionally, we will include a scatter plot graph of proportion of the population vs time for each of the three categories (race, sex, age). This will show if there was an increase or decrease in prevalence as well as a comparison between the tested variables. 

We are expecting that age will have a large effect on the mortality rate and cause an increase of numbers. Women might also have higher rates as from previous knowledge, they tend to have higher cardiovascular diseases. 

The last thing that I could add is an analysis of whether coronary heart disease had a larger or smaller influence from race, sex, or age in comparison to the influence they have on stroke. This could be shown through a two variable box plot graph of data vs. type of cardiovascular disease and display the age groups by color. 

2. For the second question of interest, we will analyze the influence that geography has on the change in prevalence of stroke and coronary heart disease has over time. We will be graphing the average change over the past 20 years in the per-capita population for stroke deaths and mortality deaths separately onto a map of the United States to see if geographic location plays a role at all in the change. Additionally, we will use a histogram in order to better evaluate the distribution of the change of stroke deaths and mortality deaths over time. 

We are not expecting state location to play a significant role in the difference between per-capita strokes and heart problems.


# Discussion
We'll discuss possible causes of various factors impacting rates of stroke and heart disease deaths. One difficult part of the results will be identifying causes vs associations. When talking about the hypothesis test results, we'll need to figure out if whether or not the factors of sex, age, and race affect the results if there appears to be a difference between the two variables. That means eliminating confounding variables. Comparing races will also be rather difficult, as there are multiple data sets so we would need to compare to the aggregate population as opposed to simply contrasting with another. 

We will also discuss the importance of the graphs and how they relate to our findings, noting distributions and outliers.

# Summary/ Conclusion
(leave blank)