---
title: "Coronary Heart Disease and Stroke Mortality Rate"
output: html_document
date: "2022-11-02"
---

Group Number: 317B

Group Member: Elli Houtakker, Nolan You, Logan Begin, Thong Nguyen, Sharvil Dave

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
library(readxl)
library(broman)
library(usmap)
source("../scripts/viridis.R")
source("../scripts/ggprob.R")
```

```{r, include = FALSE, warning = FALSE}
all_data = read_csv("data.csv")
```


```{r, include = FALSE, warning = FALSE}
##Cleaning Data
data = all_data %>%
  select(Year,LocationAbbr, LocationDesc, Topic, Data_Value, Stratification1, Stratification2, Stratification3) %>%
  rename(year = Year, state = LocationAbbr, county = LocationDesc, age_group = Stratification1, data = Data_Value, type = Topic, race = Stratification2, sex = Stratification3) %>%
  filter(!is.na(data)) %>%
  arrange(year) %>%
  mutate( type = recode(type, "Coronary heart disease (CHD)" = "CHD", "All stroke" = "Stroke"))

```


## Introduction

Strokes and heart disease are among the leading causes of death in the United States. Medical data analysis is required for research in order to figure out how to better treat patients, which is required to save lives. In this project, we are analyzing a collection of data on heart disease and stroke deaths in order to identify their roots and causes. This information is obtained directly from the CDC, which obtains it from state governments. Through this analysis, we will look at factors demographic information, like sex, age, and race, to find associations between those variables and the occurrence of heart disease and stroke. We believe that sex, age, and race are related to strokes and heart attacks. We believe there is an association between sex, age, and race and stroke and heart disease mortality rates.


## Background

In the United States, coronary heart disease is the leading cause of death for both men and women of all racial and ethnic groups, and a stroke occurs every 40 seconds^[https://www.cdc.gov/stroke/facts.htm#:~:text=Stroke%20statistics,-In%202020%2C%201&text=Every%2040%20seconds%2C%20someone%20in,minutes%2C%20someone%20dies%20of%20stroke.&text=Every%20year%2C%20more%20than%20795%2C000,are%20first%20or%20new%20strokes.].Coronary artery disease, heart attack, arrhythmia, heart failure, high blood pressure, and other heart conditions are all examples of cardiovascular disease. Many of the symptoms of heart disease go unnoticed by the patient until something major happens. Diabetes, obesity, an unhealthy diet, and a lack of physical activity are all risk factors for heart disease, but it can also be inherited.^[https://www.cdc.gov/heartdisease/facts.htm]. 
Because cardiovascular disease is a risk for everyone, many organizations study and research the effects as well as the causes in order to prevent the societal impact from continuing. The Division for Heart Disease and Stroke Prevention, Million Hearts, and WISEWOMEN are among these organizations.
^[https://www.cdc.gov/heartdisease/about.htm].


Our project will examine the relationships between stroke and coronary heart disease, as well as the relationship between their mortality rates and patient characteristics, using the data provided. The National Vital Statistics System, a division of the Centers for Disease Control and Prevention, collected this data set, which can be found directly on [CDC website](https://chronicdata.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Heart-Disease-and-Stroke-Morta/7b9s-s8ck). The Federal Government agency collects data from state medical records in order to analyze the prevalence of cardiovascular disease in the United States and make this information available to the public through the CDC website. The data collected by the CDC is used to provide information to help people make decisions to protect themselves from illness or death, as well as to provide a better understanding on diseases, prevention, and other topics. We will be analyzing county population estimates of stroke and coronary heart disease mortality rates from 1999 to 2019-collected at the end of each year-across 14 states in the United States by using this data set. The data sample is taken from the larger population which would be the entire population of the United States. The data set used a Bayesian spatiotemporal model to collect death rate trends, which is when data is collected across time and space and has at least one spatial and one temporal property.



The raw data set contains over 5 million rows and 21 columns, but we will only use 677,490 of them, excluding rows with missing information for the data value column and only selecting deaths caused by coronary heart disease (CHD) or stroke^[```{r, warning = FALSE}
# Cleaning Data
# data = all_data %>% 
#   select(Year,LocationAbbr, LocationDesc, Topic, Data_Value, Stratification1, Stratification2, Stratification3) %>% 
#   rename(year = Year, state = LocationAbbr, county = LocationDesc, age_group = Stratification1, data = Data_Value, type = Topic, race = Stratification2, sex = Stratification3) %>% 
#   filter(!is.na(data)) %>% 
#   arrange(year) %>% 
#   mutate( type = recode(type, "Coronary heart disease (CHD)" = "CHD", "All stroke" = "Stroke"))
```]. We will be interested in six variables when analyzing the data at the county level, four of which are categorical and two of which are numerical. The two numerical columns are the year the data was collected and the mortality rate, which is the proportion of the country's population of 100,000 people who died from the specified cardiovascular diseases—stroke and CHD in this case. Mortality type, age range, race, and sex are the four categorical variables. We are looking at stroke and coronary heart disease mortality rates. The age range denotes the age range in which the data was collected from females and males aged 35 to 64, as well as those aged 65 and older. In addition, the patient's race and sex are categorical variables. According to the Topic column, the numeric value represents the death rate per 100,000 people from heart disease, stroke, or heart failure.

Following the completion of preliminary research on CHD and stroke, we will analyze the data set containing mortality rates for these two causes of death in the population to answer the following two research questions:

**RQ #1:** Does age influence stroke and coronary heart disease mortality rate or its prevalence?

**RQ #2:** Does race influence stroke and coronary heart disease mortality rate or its prevalence?

**RQ #3: **Does sex influence stroke and coronary heart disease mortality rate or its prevalence?


## Analysis
### RQ1 - Age

To begin, we will examine the impact of age on the mortality rate for both stroke and CHD. The data set yielded two age groups: those aged 35 to 64 and those aged 65 and older. We will remove the rows that were not associated with a change in age; in particular, we removed the rows with the value "Overall" for age because these rows are used for analysis of other categories when age is not a factor. We obtain 51,300 rows as a result of this filtration^[```{r, warning = FALSE}
# age_data= data %>% 
#   filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease") %>% 
#   filter(race == "Overall", (sex == "Overall")) %>% 
#   select(year, state, county, type, data, age_group)
# age_data
# 
# age_sum = age_data %>% 
#   group_by(age_group, year, type) %>% 
#   mutate(c = n(),
#          n = c*100000,
#     total = sum(data)) %>% 
#   distinct(total, c,n) %>% 
#   mutate(p = total/n) 
# 
# pivot_age = age_sum %>% pivot_wider(names_from = type, values_from = c(c, total, n, p)) %>% rename(
#   n_heart = "n_CHD", 
#   p_heart = "p_CHD", 
#   p_stroke = "p_Stroke", 
#   n_stroke = "n_Stroke",
#   c_stroke = "c_Stroke", 
#   c_heart = "c_CHD",
#   total_stroke = "total_Stroke", 
#   total_heart = "total_CHD"
# )
```]. The graph below depicts this data set and shows the drastic change in proportion for both CHD and stroke deaths between both age groups, despite the fact that CHD has a higher proportion overall.

```{r, include = FALSE, warning = FALSE, echo = FALSE}
age_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease") %>% 
  filter(race == "Overall", (sex == "Overall")) %>% 
  select(year, state, county, type, data, age_group)

age_sum = age_data %>% 
  group_by(age_group, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data)) %>% 
  distinct(total, c,n) %>% 
  mutate(p = total/n) 

pivot_age = age_sum %>% pivot_wider(names_from = type, values_from = c(c, total, n, p)) %>% rename(
  n_heart = "n_CHD", 
  p_heart = "p_CHD", 
  p_stroke = "p_Stroke", 
  n_stroke = "n_Stroke",
  c_stroke = "c_Stroke", 
  c_heart = "c_CHD",
  total_stroke = "total_Stroke", 
  total_heart = "total_CHD"
)
```

```{r, warning = FALSE, echo = FALSE}
ggplot(age_sum, aes(x = year, y = p, color = type, scale= TRUE))+
  geom_point()+
  geom_smooth(se= FALSE)+
  facet_wrap(~age_group)+
  theme_minimal()+
  ggtitle("Age 35-64 vs. 65+ Mortality Rate of CHD and Stroke", subtitle = "1999 to 2019")+
  ylab("Proportion of Mortality by Disease per 100,000 population")+
  xlab("Year")+
  theme(axis.title.y = element_text(size = 9)) +
  theme(axis.title.x = element_text(size = 12)) 


```


#### Hypothesis Tests:

Using the data shown above, we will now run the following hypothesis test. This test will determine whether the average morality rate for all counties, races, and genders in the age group of 35 to 64 is equal to or not equal to the average morality rate for those aged 65 and older under the same conditions. To control the influence on death and compare the two causes of death, we will test the hypothesis separately for each cause of death, stroke and CHD. Based on the central limit theorem, we will assume that the mean distribution is approximately normal for each of the hypothesis tests. The central limit theorem assumes that as sample size increases to infinity, the exact probability distribution of a sample mean converges to a normal distribution.

**Coronary Heart Disease Hypothesis Tests:**

- We will assume that there is a universal parameters $\mu_{\text{35-64}}$ and $\mu_{\text{65+}}$  representing the mean of mortality due to coronary heart disease  per year under the experimental conditions aged 35 - 64 and those aged 65 and older, respectively. 
- Assume the distribution of the mean of mortality due to coronary heart disease is approximately normal
- Assume that all trials are independent.

Model:

$X_i \sim N(\mu_{\text{35-64}}, \sigma_{\text{35-64}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim N(\mu_{\text{65+}}, \sigma_{\text{65+}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{35-64}} = \mu_{\text{65+}}$    
$H_a: \mu_{\text{35-64}} \neq \mu_{\text{65+}}$

Test statistic:

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r, warning = FALSE, echo = FALSE}

x = age_data %>% filter(type == "CHD" & age_group == "Ages 35-64 years") %>% pull(data)
y = age_data %>% filter(type == "CHD" & age_group != "Ages 35-64 years") %>% pull(data)
age_chd_test = t.test(x, y)
age_chd_test
```
There is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from coronary heart disease is smaller than the mean of the mortality rate of those aged 65 and older from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances).


**Stroke Hypothesis Tests:**

- We will assume that there is a universal parameters $\mu_{\text{35-64}}$ and $\mu_{\text{65+}}$  representing the mean of mortality due to stroke per year under the experimental conditions aged 35 - 64 and those aged 65 and older, respectively. 
- Assume the distribution of the mean of mortality due to stroke is approximately normal
- Assume that all trials are independent.

Model:

$X_i \sim N(\mu_{\text{35-64}}, \sigma_{\text{35-64}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim N(\mu_{\text{65+}}, \sigma_{\text{65+}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{35-64}} = \mu_{\text{65+}}$    
$H_a: \mu_{\text{35-64}} \neq \mu_{\text{65+}}$

Test statistic

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r, warning = FALSE, echo = FALSE}

x = age_data %>% filter(type == "Stroke" & age_group == "Ages 35-64 years") %>% pull(data)
xy = age_data %>% filter(type == "Stroke" & age_group != "Ages 35-64 years") %>% pull(data)
age_stroke_test = t.test(x, y)
age_stroke_test
```
There is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from stroke is smaller than the mean of the mortality rate of those aged 65 and older from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances).

**Age Influence on Heart Disease vs. Stroke:**

We will now compare the difference in mortality rate per 100,000 population between the two deaths. Through the graph, it is shown that the average mortality rate for CHD is higher among those aged 65 and older compared to those who died from stroke. The average is comparable for the people aged 35 to 64, though. 

```{r, warning = FALSE, echo = FALSE}
ggplot(age_data, aes(y = data, x = type,  fill = age_group)) +
  geom_boxplot()+
  ggtitle("CHD and Stroke Mortality by Age")+
  xlab("Cause of Death")+
  ylab("Mortality per 100,000 population")
```


### RQ2 - Sex

To address RQ2, we will evaluated the effect of sex on mortality rates for each cause of death, stroke and CHD, and see if the associations are similar between the two. We will remove the rows that were not associated with a change in sex; specifically, we removed the rows with the sex value "Overall" because these rows are used for analyzing other categories when sex is not a factor. As a result of this filtration^[```{r, warning = FALSE}
# sex_data= data %>% 
#   filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease", race == "Overall", sex != "Overall") %>% 
#   select(!race)
# sex_data
# 
# sex_sum = sex_data %>% 
#   group_by(sex, year, type) %>% 
#   mutate(c = n(),
#          n = c*100000,
#     total = sum(data)) %>% 
#   distinct(total, c,n) %>% 
#   mutate(p = total/n)
```], we get 99,557 rows, which are represented by the graph below. These two scatter plots show that CHD is a more common cause of death than stroke and that there is a significant difference between men and women from 1999 to 2021. Meanwhile, strokes kill nearly the same number of men and women across the data set. We will put these observations to the test further with the hypothesis test that follows.


```{r, include = FALSE, warning = FALSE}
sex_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease", race == "Overall", sex != "Overall") %>% 
  select(!race)

sex_sum = sex_data %>% 
  group_by(sex, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data)) %>% 
  distinct(total, c,n) %>% 
  mutate(p = total/n)
```

```{r, warning = FALSE, echo = FALSE}
ggplot(sex_sum, aes(x = year, y = p, color = type, scale= TRUE))+
  geom_point()+
  geom_smooth(se= FALSE)+
  facet_wrap(~sex)+
  theme_minimal()+
  ggtitle("Mortality Rate of CHD and Stroke by Sex", subtitle = "1999-2019")+
  ylab("Proportion of Mortality by Disease per 100,000 population")+
  xlab("Year")+
  theme(axis.title.y = element_text(size = 9)) +
  theme(axis.title.x = element_text(size = 12)) 
```


#### Hypothesis Tests:
This hypothesis test will determine whether the average mortality rate is the same for both sexes (male and female) or if they differ. To control the influence on death, we test the hypothesis for each cause of death separately. The assumptions are similar to the tests done in the first steps, under the same condition of the central limit theorem.

**Coronary Heart Disease Hypothesis Tests:**


- We will assume that there is a universal parameters $\mu_{\text{female}}$ and $\mu_{\text{male}}$  representing the mean of female and male mortality due to coronary heart disease per year, respectively. 
- Assume that all trials are independent
- Assume the distribution of the mean of mortality due to coronary heart disease is approximately normal


Model:

$X_i \sim N(\mu_{\text{female}}, \sigma_{\text{female}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim N(\mu_{\text{male}}, \sigma_{\text{male}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{female}} = \mu_{\text{male}}$    
$H_a: \mu_{\text{female}} \neq \mu_{\text{male}}$

Test statistic:

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r, warning = FALSE, echo = FALSE}

x = sex_data %>% filter(type == "CHD" & sex == "Women") %>% pull(data)
y = sex_data %>% filter(type == "CHD" & sex == "Men") %>% pull(data)
sex_chd_test = t.test(x, y)
sex_chd_test
```
There is very strong evidence that the mean mortality rate of women who died from coronary heart disease is smaller than the mean of the mortality rate of men from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances, extreme test statistic).


**Stroke Hypothesis Tests:**

- We will assume the same assumptions and definition of the $\mu_{\text{female}}$ and $\mu_{\text{male}}$ parameters as the hypothesis above.

Model:

$X_i \sim N(\mu_{\text{female}}, \sigma_{\text{female}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim N(\mu_{\text{male}}, \sigma_{\text{male}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{female}} = \mu_{\text{male}}$    
$H_a: \mu_{\text{female}} \neq \mu_{\text{male}}$

Test statistic

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r, warning = FALSE, echo = FALSE}

x = sex_data %>% filter(type == "Stroke" & sex == "Women") %>% pull(data)
y = sex_data %>% filter(type == "Stroke" & sex == "Men") %>% pull(data)
sex_stroke_test = t.test(x, y)
sex_stroke_test
```

There is not evidence to conclude that the mean mortality rate from stroke is different between men and women.(High p value, two-sided t-test, test statistic close to 0).

**Sex Influence on Heart Disease vs. Stroke:**

The difference in mortality rate per 100,000 population between the two deaths by age is shown through the graph. The average mortality is influenced more by sex for coronary heart disease compared to stroke deaths. Between sexes for stroke data, it has almost identical statistical data analysis while CHD differs in average, and upper and lower quantitles. 

```{r, warning = FALSE, echo = FALSE}
ggplot(sex_data, aes(y = data, x = type,  fill = sex)) +
  geom_boxplot()+
  ggtitle("CHD and Stroke Mortality by Sex")+
  xlab("Cause of Death")+
  ylab("Mortality per 100,000 Population")
```


### RQ3 - Race

The mortality rate from coronary heart disease and stroke may also be affected by race. To answer RQ3, we will see if there is a difference between a specific race and the overall average of all races. We will remove the rows that were not associated with a change in race; in particular, we removed the rows with the race value "Overall" because these rows are used to analyze other categories when race is not a factor^[```{r, warning = FALSE, include = FALSE}
# race_data= data %>% 
#   filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease", race != "Overall", sex == "Overall") %>% 
#   select(!sex)
# 
# race_sum = race_data %>% 
#   group_by(race, year, type) %>% 
#   mutate(c = n(),
#          n = c*100000,
#     total = sum(data),
#     mean = mean(data)) %>% 
#   distinct(total, c,n, mean) %>% 
#   mutate(p = total/n)
# race_sum
```]. As a result, this graph set contains 120,145 rows. The red dashed line represents the average mortality rate across all races. This indicates how far each race differs from the overall average if race was not a factor.

```{r, include = FALSE, warning = FALSE, echo = FALSE}
race_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease", race != "Overall", sex == "Overall") %>% 
  select(!sex)

race_sum = race_data %>% 
  group_by(race, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data),
    mean = mean(data)) %>% 
  distinct(total, c,n, mean) %>% 
  mutate(p = total/n)
race_sum
```

```{r, warning = FALSE, echo = FALSE}
mean= race_data %>% summarize(mean = mean(data)) %>% pull(mean)
ggplot(race_sum, aes(x = year, y = mean, color = type, scale= TRUE))+
  geom_point()+
  geom_smooth(se= FALSE)+
  facet_wrap(~race)+
  theme_minimal()+
  geom_hline(yintercept =mean, linetype = "dashed", color = "red")+
  ggtitle("Average Mortality of CHD and Stroke by Race", subtitle = "1999-2019")+
  ylab("Average Mortality per 100,000 population")+
  xlab("Year")+
  theme(axis.title.y = element_text(size = 9)) +
  theme(axis.title.x = element_text(size = 12)) 
```

#### Hypothesis Tests: 

These two hypothesis tests will be used to determine whether the average mortality rate is the same between races (male and female) or if they do not equal each other. The following races are represented in the data set: Black (non-Hispanic), Hispanic, White, American Indian/Alaska Native, and Asian/Pacific Islander. To control the influence on death, we test the hypothesis for each cause of death separately. We will keep the same assumptions as above that are defined under the central limit theorem.

**Coronary Heart Disease Hypothesis Tests:**


- We will assume that there is a universal parameters $\mu_{\text{all other}}$ and $\mu_{\text{black}}$ representing the mean mortality due to coronary heart disease per year over all races, sex, and age, and the mean mortality due to coronary heart disease of Asians/Pacific Islanders, respectively. 
- Assume that all trials are independent.

Model:

$X_i \sim N(\mu_{\text{all other}}, \sigma_{\text{all other}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim N(\mu_{\text{black}}, \sigma_{\text{black}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{all other}} = \mu_{\text{black}}$    
$H_a: \mu_{\text{all other}} < \mu_{\text{black}}$

Test statistic:

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r, warning = FALSE, echo = FALSE}
x = race_data %>% filter(type == "CHD" & race != "Black (Non-Hispanic)") %>% pull(data)
y = race_data %>% filter(type == "CHD" & race == "Black (Non-Hispanic)") %>% pull(data)
race_chd_test = t.test(x, y, alternative = c("less"))
race_chd_test
```
	
There is strong statistically evidence that the mean mortality rate of Non-Hispanic Black people is disproportionately higher than the average of all other races. There is a difference between the groups, and we can see that by its low p value(p-value = 2.2e-16, two sided t-test) and extreme test statistic.


**Stroke Hypothesis Tests:**

- We will assume that there is a universal parameters $\mu_{\text{all other}}$ and $\mu_{\text{asian}}$ representing the mean mortality due to stroke per year over all races, sex, and age, and the mean mortality due to stroke of Blacks(non Hispanics), respectively. 
- Assume that all trials are independent.

Model:

$X_i \sim N(\mu_{\text{all other}}, \sigma_{\text{all other}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim N(\mu_{\text{asian}}, \sigma_{\text{asian}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{all other}} = \mu_{\text{asian}}$    
$H_a: \mu_{\text{all other}} > \mu_{\text{asian}}$

Test statistic:

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r, warning = FALSE, echo = FALSE}

x = race_data %>% filter(type == "Stroke" & race != "Asian/Pacific Islander") %>% pull(data)
y = race_data %>% filter(type == "Stroke" & race == "Asian/Pacific Islander") %>% pull(data)
race_stroke_test = t.test(x, y, alternative = c("greater"))
race_stroke_test
```

There is strong evidence that the mean mortality rate of Asians is disproportionately lower than the average mortality rate of all other races. There is a difference between the groups, and we can see that by its low p value(p-value = 2.2e-16, two sided t-test) and extreme test statistic.


**Race Influence on Heart Disease vs. Stroke:**

Overall, there seems to be trend between the higher influence of race in those who suffered from CHD compared to those of stroke. This is demonstrated through the higher range quantities for CHD and the higher differences between each race. Stroke moralities are all under 1500 and have an average close to 115. 

```{r, warning = FALSE, echo = FALSE}
ggplot(race_data, aes(y = data, x = type,  fill = race)) +
  geom_boxplot()+
  ggtitle("CHD and Stroke Mortality by Race")+
  xlab("Cause of Death") +
  ylab("Mortality per 100,000")
```

```{r, warning = FALSE, echo = FALSE}
# age_data2 = age_data %>% group_by(year, age_group, type) %>% 
#   summarize(data = sum(data)/n())
# race_data2 = race_data %>% group_by(year, race, type) %>% 
#   summarize(data = sum(data)/n())
# sex_data2 = sex_data %>% group_by(year, sex, type) %>% 
#   summarize(data = sum(data)/n())
# ggplot()+
#   geom_smooth(data = age_data2, aes(x = year, y = data), color = "red", se = FALSE)+
#   geom_smooth(data = race_data2, aes(x = year, y = data), color = "black", se = FALSE)+
#   geom_smooth(data = sex_data2, aes(x = year, y = data), color = "purple", se = FALSE)+
#   facet_grid(~type)+
#   ggtitle("Fitted Line for CHD")+
#   xlab("Years")+
#   ylab("Average Mortality Rate")
# ggplot()+
#   geom_density(data = age_data2, aes( x = data), fill = "red", alpha = 0.5)+
#   geom_density(data = race_data2, aes(x = data), fill = "black", alpha = 0.5)+
#   geom_density(data = sex_data2, aes( x = data), fill = "purple",alpha = 0.5)+
#   facet_grid(~type)+
#   ggtitle("Density Plot for CHD")+
#   xlab("Average Mortality Rate")+
#   ylab("Density")

```




### Analysis Summary

In conclusion, we can divide our result into 3 categories including Age, Sex, and Race:

- **Age:** There is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from coronary heart disease is different from the mean of the mortality rate of those aged 65 and older from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances). Also, there is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from stroke is different from the mean of the mortality rate of those aged 65 and older from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances).

- **Sex:**There is very strong evidence that the mean mortality rate of women who died from coronary heart disease is smaller than the mean of the mortality rate of men from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances, extreme test statistic). Also, there is not evidence to conclude that the mean mortality rate from stroke is different between men and women.(High p value, two-sided t-test, test statistic close to 0).

- **Race:** There is strong statistically evidence that the mean mortality rate of Non-Hispanic Black people is disproportionately higher than the average of all other races. There is a difference between the groups, and we can see that by its low p value(p-value = 2.2e-16, two sided t-test) and extreme test statistic. Also, there is strong evidence that the mean mortality rate of Asians is disproportionately lower than the average mortality rate of all other races. There is a difference between the groups, and we can see that by its low p value(p-value = 2.2e-16, two sided t-test) and extreme test statistic.


**Stroke Analysis:**
```{r, echo = FALSE, warning = FALSE}
stroke_analysis = tibble("Characteristic" = c("Age", "Sex", "Race"), "Hypothesis Test" = c("H_a: mu{35-64} != mu{65+}", "H_a: mu{female} != mu{male}", "H_a: mu{all other} > mu{asian}"), "X Mean" = c(20.75, 178.93, 161.72), "Y Mean" = c( 791.88, 179.08, 112.43), "Test Statistic" = c( -293.92, -0.098, 27.92), "P-Value" = c(2.2e-16,0.92, 2.2e-16))
stroke_analysis
```

**Coronary Heart Disease Analysis:**
```{r, echo = FALSE, warning = FALSE}
chd_analysis = tibble("Characteristic" = c("Age", "Sex", "Race"), "Hypothesis Test" = c("H_a: mu{35-64} != mu{65+}", "H_a: mu{female} != mu{male}", "H_a: mu{all other} > mu{black}"), "X Mean" = c(77.73, 330.96, 335.37), "Y Mean" = c( 791.88, 556.32, 399.81), "Test Statistic" = c( -270.64, -57.121, -16.225), "P-Value" = c(2.2e-16,2.2e-16, 2.2e-16))
chd_analysis
```



# Discussion


Based on our data analysis, we can say that the factors of age, race and sex are each associated with CHD and stroke mortality rate. This is because there is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from coronary heart disease is different from the mean of the mortality rate of those aged 65 and older. Also, there is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from stroke is different from the mean of the mortality rate of those aged 65 and older. There is very strong evidence that the mean mortality rate of women who died from coronary heart disease is smaller than the mean of the mortality rate of men from heart disease. Also, there is not evidence to conclude that the mean mortality rate from stroke is different between men and women. There is strong statistically evidence that the mean mortality rate of Non-Hispanic Black people is disproportionately higher than the average of all other races. There is a difference between the racial groups, and we can see that by its low p value and extreme test statistic. Also, there is strong evidence that the mean mortality rate of Asians is disproportionately lower than the average mortality rate of all other races. There is a difference between the racial groups, and we can see that by its low p value and extreme test statistic. 


There are three shortcomings in our project. First, we are not able to create a generalization for the entire U.S. because certain states are not included in the data, which could be solved if we had data from every U.S. state. Additionally, we may have drawn conclusions predicated on faulty assumptions. For example, such as the reporting of who dies from a CHD or a stroke might actually be biased and non-indicative of the overall parent population of everyone, not random, so we cannot assume that the data is random. 

Additionally, another research question that could have been investigated further using this data set in conjunction with another is how these mortality rates compare to factors such as smoking, obesity rate, and others. In essence, the study would seek another variable that would be associated with the change in CHD and stroke mortality rates over time and refine our understanding of the cause of death. In order to properly depict and draw conclusions from this data, we would need to collect another data set, such as a full medical record or similar, on these factors that is controlled by the same variables. This could have greatly benefited the expansion of our research questions by including history of smoking, prediagnosed conditions, or whether or not that disease is hereditary. 


