---
title: "Draft Project"
output: html_document
date: "2022-11-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
library(readxl)
source("../scripts/viridis.R")
source("../scripts/ggprob.R")
```


### Coronary Heart Disease and Stroke

```{r}
#all_data = read_excel("heart_disease_data.xlsx", sheet = 1)
all_data = read_csv("data.csv")
all_data
```


```{r}
##Cleaning Data
data = all_data %>% 
  select(Year,LocationAbbr, LocationDesc, Topic, Data_Value, Stratification1, Stratification2, Stratification3) %>% 
  rename(year = Year, state = LocationAbbr, county = LocationDesc, age_group = Stratification1, data = Data_Value, type = Topic, race = Stratification2, sex = Stratification3) %>% 
  filter(!is.na(data)) %>% 
  arrange(year) %>% 
  mutate( type = recode(type, "Coronary heart disease (CHD)" = "CHD", "All stroke" = "Stroke"))
data

```


# Introduction



# Background

Our question is whether age, race, and gender influence stroke and heart disease mortality, and whether geographical location influences the prevalence of these cardiovascular diseases.

Coronary heart disease is the leading cause of death in the United States for both men and women of all racial and ethnic groups, and a stroke occurs every 40 seconds. As a result, analyzing cardiovascular disease is common for many Americans and can affect anyone. Many organizations study and research both the effects and the causes in order to prevent the impact on society from continuing, but there is a general lack of funding for many health research facilities. The Division for Heart Disease and Stroke Prevention, Million Hearts, and WISEWOMEN are a few of these organizations.

Cardiovascular disease encompasses a wide range of heart conditions such as coronary artery disease, heart attack, arrhythmia, heart failure, high blood pressure, and others. Many of the symptoms of heart disease are not noticed by the patient until a major event occurs. Diabetes, obesity, an unhealthy diet, and lack of physical activity all increase one's risk of heart disease, but it can also be inherited.


Our project will use the provided data to examine the relationships between stroke and coronary heart disease, as well as patient characteristics and geographic location. This data set was collected by the National Vital Statistics System, a division of the Centers for Disease Control and Prevention, and can be found directly on the [CDC website](https://chronicdata.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Heart-Disease-and-Stroke-Morta/7b9s-s8ck). This Federal Government agency obtains its information from state medical records. The data set includes population estimates of stroke and coronary heart disease mortality rates from 1999 to 2019. The data set used a Bayesian spatiotemporal model that was smoothed over time, space, and age group to collect death rate trends.


There are over 5 million rows and 21 columns in the raw data set. We will be interested in eight variables, six of which are categorical and two of which are numerical and contain the estimated death rate value and the year data was collected. The four categorical variables are State, County, cardiovascular disease, age range, race, and gender. The state and county will be used to determine the geographical dependencies of the data values, as well as which counties have a higher prevalence. The age range indicates where the data was collected from females and males aged 35 to 64, as well as those aged 65 and older. There are also two categorical variables that contain the data's race and gender. The numeric value represents the death rate per 100,000 people from heart disease, stroke, or heart failure, as specified by the Topic column.


Cites: https://www.cdc.gov/stroke/facts.htm#:~:text=Stroke%20statistics,-In%202020%2C%201&text=Every%2040%20seconds%2C%20someone%20in,minutes%2C%20someone%20dies%20of%20stroke.&text=Every%20year%2C%20more%20than%20795%2C000,are%20first%20or%20new%20strokes.
https://www.cdc.gov/heartdisease/facts.htm
https://www.cdc.gov/heartdisease/about.htm



# Analysis
## Age

```{r}
age_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease") %>% 
  filter(race == "Overall", (sex == "Overall")) %>% 
  select(year, state, county, type, data, age_group)
age_data

age_sum = age_data %>% 
  group_by(age_group, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data)) %>% 
  distinct(total, c,n) %>% 
  mutate(p = total/n) 
age_sum

pivot_age = age_sum %>% pivot_wider(names_from = type, values_from = c(c, total, n, p)) %>% rename(
  n_heart = "n_CHD", 
  p_heart = "p_CHD", 
  p_stroke = "p_Stroke", 
  n_stroke = "n_Stroke",
  c_stroke = "c_Stroke", 
  c_heart = "c_CHD",
  total_stroke = "total_Stroke", 
  total_heart = "total_CHD"
)
pivot_age
```

```{r}
ggplot(age_sum, aes(x = year, y = p, color = type, scale= TRUE))+
  geom_point()+
  geom_smooth(se= FALSE)+
  facet_wrap(~age_group)+
  theme_bw()+
  ggtitle("Scatter Plot of Mortality Rate by Age", subtitle = "grouped by Cardiovascular type")+
  ylab("Mortality Proportion of Population")+
  xlab("Year")

```


### Coronary Heart Disease Hypothesis Tests:

- We will assume that there is a universal parameters $\mu_{\text{35-64}}$ and $\mu_{\text{65+}}$  representing the mean of mortality due to coronary heart disease  per year under the experimental conditions aged 35 - 64 and those aged 65 and older, respectively. 
- Assume the distribution of the mean of mortality due to coronary heart disease is approximately normal (We are interested in the sampling distribution of the sample)
- Assume that all trials are independent.

Model:

$X_i \sim F_1(\mu_{\text{35-64}}, \sigma_{\text{35-64}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim F_2(\mu_{\text{65+}}, \sigma_{\text{65+}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{35-64}} = \mu_{\text{65+}}$    
$H_a: \mu_{\text{35-64}} \neq \mu_{\text{65+}}$

Test statistic:

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r}

x = age_data %>% filter(type == "CHD" & age_group == "Ages 35-64 years") %>% pull(data)
y = age_data %>% filter(type == "CHD" & age_group != "Ages 35-64 years") %>% pull(data)
t.test(x, y)
```
There is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from coronary heart disease is smaller than the mean of the mortality rate of those aged 65 and older from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances).


### Stroke Hypothesis Tests:

- We will assume that there is a universal parameters $\mu_{\text{35-64}}$ and $\mu_{\text{65+}}$  representing the mean of mortality due to stroke per year under the experimental conditions aged 35 - 64 and those aged 65 and older, respectively. 
Assume the distribution of the mean of mortality due to stroke is approximately normal (We are interested in the sampling distribution of the sample)
- Assume that all trials are independent.

Model:

$X_i \sim F_1(\mu_{\text{35-64}}, \sigma_{\text{35-64}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim F_2(\mu_{\text{65+}}, \sigma_{\text{65+}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{35-64}} = \mu_{\text{65+}}$    
$H_a: \mu_{\text{35-64}} \neq \mu_{\text{65+}}$

Test statistic

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r}

x = age_data %>% filter(type == "Stroke" & age_group == "Ages 35-64 years") %>% pull(data)
xy = age_data %>% filter(type == "Stroke" & age_group != "Ages 35-64 years") %>% pull(data)
t.test(x, y)
```
There is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from stroke is smaller than the mean of the mortality rate of those aged 65 and older from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances).

### Age Influence on Heart Disease vs. Stroke:
```{r}
age_data %>% 
  group_by(type, age_group) %>% 
  summarize(mean = mean(data),
            q10 = quantile(data, 0.10),
            q25 = quantile(data, 0.25),
            q50 = quantile(data, 0.5),
            q75 = quantile(data, 0.75),
            q90 = quantile(data, 0.9))
```

```{r}
ggplot(age_data, aes(y = data, x = type,  fill = age_group)) +
  geom_boxplot()
```


## Sex

```{r}
sex_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease", race == "Overall", sex != "Overall") %>% 
  select(!race)
sex_data

sex_sum = sex_data %>% 
  group_by(sex, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data)) %>% 
  distinct(total, c,n) %>% 
  mutate(p = total/n)
```

```{r}
ggplot(sex_sum, aes(x = year, y = p, color = type, scale= TRUE))+
  geom_point()+
  geom_smooth(se= FALSE)+
  facet_wrap(~sex)+
  theme_bw()+
  ggtitle("Scatter Plot of Mortality Rate by Sex", subtitle = "grouped by Cardiovascular type")+
   ylab("Mortality Proportion of Population")+
  xlab("Year")
```


### Coronary Heart Disease Hypothesis Tests:

- We will assume that there is a universal parameters $\mu_{\text{female}}$ and $\mu_{\text{male}}$  representing the mean of female and male mortality due to coronary heart disease per year, respectively. 
- Assume that all trials are independent.

Model:

$X_i \sim F_1(\mu_{\text{female}}, \sigma_{\text{female}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim F_2(\mu_{\text{male}}, \sigma_{\text{male}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{female}} = \mu_{\text{male}}$    
$H_a: \mu_{\text{female}} \neq \mu_{\text{male}}$

Test statistic:

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r}

x = sex_data %>% filter(type == "CHD" & sex == "Women") %>% pull(data)
y = sex_data %>% filter(type == "CHD" & sex == "Men") %>% pull(data)
t.test(x, y)
```
***There is very strong evidence that the mean mortality rate of male and female from heart disease(CHD) are different from each other ($p = 2.2e-16$, two-sided t-test, unequal variances).**** EDIT


### Stroke Hypothesis Tests:

- We will assume the same assumptions and definition of the $\mu_{\text{female}}$ and $\mu_{\text{male}}$ parameters as the hypothesis above.

Model:

$X_i \sim F_1(\mu_{\text{female}}, \sigma_{\text{female}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim F_2(\mu_{\text{male}}, \sigma_{\text{male}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{female}} = \mu_{\text{male}}$    
$H_a: \mu_{\text{female}} \neq \mu_{\text{male}}$

Test statistic

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r}

x = sex_data %>% filter(type == "Stroke" & sex == "Women") %>% pull(data)
y = sex_data %>% filter(type == "Stroke" & sex == "Men") %>% pull(data)
t.test(x, y)
```

***There is very strong evidence that the mean mortality rate of stroke of male and female are the same ($p = 0.922$, two-sided t-test, unequal variances).*** 

### Sex Influence on Heart Disease vs. Stroke:

```{r}
sex_data %>% 
  group_by(type, sex) %>% 
  summarize(mean = mean(data),
            q10 = quantile(data, 0.10),
            q25 = quantile(data, 0.25),
            q50 = quantile(data, 0.5),
            q75 = quantile(data, 0.75),
            q90 = quantile(data, 0.9))
```

```{r}
ggplot(sex_data, aes(y = data, x = type,  fill = sex)) +
  geom_boxplot()
```


## Race

```{r}
race_data= data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease", race != "Overall", sex == "Overall") %>% 
  select(!sex)

race_sum = race_data %>% 
  group_by(race, year, type) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data),
    mean = mean(data)) %>% 
  distinct(total, c,n, mean) %>% 
  mutate(p = total/n)
race_sum
```

```{r}
mean= race_data %>% summarize(mean = mean(data)) %>% pull(mean)
ggplot(race_sum, aes(x = year, y = mean, color = type, scale= TRUE))+
  geom_point()+
  geom_smooth(se= FALSE)+
  facet_wrap(~race)+
  theme_bw()+
  geom_hline(yintercept =mean, linetype = "dashed", color = "red")+
  ggtitle("Scatter Plot of Mortality Rate by Race", subtitle = "grouped by Cardiovascular type")+
  ylab("Mortality Proportion of population")+
  xlab("Year")
```

### Coronary Heart Disease Hypothesis Tests:

- We will assume that there is a universal parameters $\mu_{\text{all}}$ and $\mu_{\text{black}}$ representing the mean mortality due to coronary heart disease per year over all races, sex, and age, and the mean mortality due to coronary heart disease of Asians/Pacific Islanders, respectively. 
- Assume that all trials are independent.

Model:

$X_i \sim F_1(\mu_{\text{all}}, \sigma_{\text{all}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim F_2(\mu_{\text{black}}, \sigma_{\text{black}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{all}} = \mu_{\text{black}}$    
$H_a: \mu_{\text{all}} \neq \mu_{\text{black}}$

Test statistic:

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r}
x = race_data %>% filter(type == "CHD") %>% pull(data)
y = race_data %>% filter(type == "CHD" & race == "Black (Non-Hispanic)") %>% pull(data)
t.test(x, y)
```
***There is very strong evidence that the mean mortality rate of black (non-Hispanic) people and that of average people are the same ($p = 2.2e-16$, two-sided t-test, unequal variances).***


### Stroke Hypothesis Tests:

- We will assume that there is a universal parameters $\mu_{\text{all}}$ and $\mu_{\text{asian}}$ representing the mean mortality due to stroke per year over all races, sex, and age, and the mean mortality due to stroke of Blacks(non Hispanics), respectively. 
- Assume that all trials are independent.

Model:

$X_i \sim F_1(\mu_{\text{all}}, \sigma_{\text{all}}), \quad i = 1, \ldots, n_1$    
$Y_i \sim F_2(\mu_{\text{asian}}, \sigma_{\text{asian}}), \quad i = 1, \ldots, n_2$   

Hypotheses:

$H_0: \mu_{\text{all}} = \mu_{\text{asian}}$    
$H_a: \mu_{\text{all}} \neq \mu_{\text{asian}}$

Test statistic:

$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r}

x = race_data %>% filter(type == "Stroke") %>% pull(data)
y = race_data %>% filter(type == "Stroke" & race == "Asian/Pacific Islander") %>% pull(data)
t.test(x, y)
```

***There is very strong evidence that the mean mortality rate of Asian people and that of average people are the same ($p = 2.2e-16$, two-sided t-test, unequal variances)***


### Race Influence on Heart Disease vs. Stroke:
```{r}
ggplot(race_data, aes(y = data, x = type,  fill = race)) +
  geom_boxplot()
```

## Geography

# Discussion

```{r}
dis_data = data %>% 
  filter(type != "Heart failure" & type != 	"Cardiovascular disease (CVD)" & type != "All heart disease") %>% 
  group_by(year,type, sex, race) %>% 
  mutate(c = n(),
         n = c*100000,
    total = sum(data)) %>% 
  distinct(total, c, n) %>% 
  mutate(p = total/n)
dis_data

proportion_stroke = dis_data %>% 
  group_by(type) %>% 
  summarise(mean_p = mean(p)) 
proportion_stroke
```
Beside surveying the mean number of people getting CHD and stroke, we can have some more questions about what is exactly the proportion of getting these diseases. In the following observation, we surveyed whether Black (non-Hispanic) people have higher rate of getting CHD and stroke. The average proportion of getting CHD among sample population in recent decades is 0.0035 and the average proportion of getting Stroke is 0.00152. Use a hypothesis test to test if the average proportion of Black (Non-Hispanic) getting CHD and stroke among sample population in recent decades are equal to 0.0035 and 0.00152 respectively versus the alternative that they are different.

$$
H_0: p_{\text{disease of black}}  = p_{\text{disease of all}} \\
H_a: p_{\text{disease of black}}  > p_{\text{disease of all}}
$$

```{r}
bl_df = dis_data %>% 
  filter (race == "Black (Non-Hispanic)")

stroke_df =  bl_df %>% filter(type == "Stroke")
chd_df = bl_df %>% filter(type == "CHD")

x_stroke = stroke_df %>%  pull(p)
x_CHD = chd_df %>%  pull(p)

t.test(x_stroke)
t.test(x_CHD)

blackstroke_sum=  bl_df %>% 
  group_by(type) %>% 
  summarise(n = n(),
            xbar = mean(p),
            sd = sd(p),
            se = sd / sqrt(n))
blackstroke_sum
```
> We are 95% confident that the mean proportion of Black (Non-hispanic) people getting stroke would have been between 0.001851512 and 0.002258730.
> We are 95% confident that the mean proportion of Black (Non-hispanic) people getting CHD would have been between 0.003515034 and 0.004481209.

A plot representing the density of proportion of getting these diseases

```{r}
ggplot(stroke_df, aes(x = p))+
  geom_density(fill = 'Orange')+
  ggtitle("Density of the proportion of getting stroke of Black (Non-hispanic) people") +
  xlab("Proportion")+
  geom_hline(yintercept = 0)

ggplot(chd_df, aes(x = p))+
  geom_density(fill = 'Red') +
  ggtitle("Density of the proportion of getting CHD of Black (Non-hispanic) people")+
  xlab("Proportion")+
  geom_hline(yintercept = 0)
```

```{r}
mean0_p_stroke = 0.00152
mean0_p_chd = 0.0035
tstat_stroke  = blackstroke_sum %>% 
  filter (type == 'Stroke') %>% 
  mutate( tstat = (xbar -  mean0_p_stroke)/se,
          pval_stroke = 1- pt(tstat, n-1))
tstat_stroke

tstat_s = tstat_stroke$tstat

tstat_chd  = blackstroke_sum %>% 
  filter (type == 'CHD') %>% 
  mutate( tstat = (xbar -  mean0_p_chd)/se,
          pval_stroke = 1- pt(tstat, n-1))
tstat_chd

tstat_c = tstat_chd$tstat
```

```{r}
gt(df = 20)+
  geom_vline(xintercept = tstat_s, linetype = "dashed", colour = 'red') +
  geom_vline(xintercept = tstat_c, linetype = "dashed", colour = 'black')
```

> Conclusion: There is very strong evidence (p= 0.022, and p = 1.145968e-05, respectively, one-sided t-test) that the average proportion of Black (Non-hispanic) people has higher chance to get stroke and CHD than those of overall people, which is 0.00152 and 0.0035, respectively.

In addition, we have the similar questions around the data: whether Asian people have higher proportion of getting diseases than the average of the overall, how much are the proportions, whether they are consistent throughout the years, etc. However, our models might be based on fault assumptions:
- The data collected is biased, not random, so we cannot assume these data is random, and making models for the larger population.
- The data might not be continuous (There might not be possible the number of people getting disease).
- The distribution might not be approximately normal.

For further analyzing, we can dig deeper into factors affecting one group having higher average chance of getting disease (Stroke and CHD), making regression to predict the pattern of the average proportion of people getting disease in different regions , etc.

In conclusion, we devide our result into 3 categories including Age, Sex, and Race:
- Age: There is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from coronary heart disease is different from the mean of the mortality rate of those aged 65 and older from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances). Also, there is very strong evidence that the mean mortality rate of those aged 35 to 64 who died from stroke is different from the mean of the mortality rate of those aged 65 and older from heart disease ($p = 2.2e-16$, two-sided t-test, unequal variances).
- Sex:
- Race:


# Analysis Summary


# References
